Testing for measurement invariance with respect to an ordinal variable

### 2013-04-25 ordinal/ordinal.Rnw
Minor changes to ordinal ms to remove color from figures.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@204 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

If $V$ is only nominal/categorical,
there is not even a partial ordering, i.e., measurement invariance tests should neither
exploit the ordering of $V$'s levels nor of the observations within the level.
In this situation, it is possible to obtain a test statistic by first summing scores within
each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where the first subscript on the two terms in parentheses are $i_\ell$
and $i_{\ell-1}$, respectively.  Again, tests of subsets of model
parameters can be obtained by 
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and likelihood ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed above explicitly account for the fact
that $V$ is ordinal.

@@

plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
     xlab = "Age group", ylab = "Weighted max statistics")
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)

plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
     xlab = "Age group", ylab = "LM statistics")
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)
@
\end{figure}

@@

plot(gefp2, functional = ordwmax(gefp2), axes = FALSE, main = "",
     xlab = "Age group", ylab = "Weighted max statistics")
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)

plot(gefp2, functional = ol2bb2.fun, axes = FALSE, ylim = c(0, 115), main = "",
     xlab = "Age group", ylab = "LM statistics")
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)
@
\end{figure}

>>>>

If $V$ is only nominal/categorical,
there is not even a partial ordering, i.e., measurement invariance tests should neither
exploit the ordering of $V$'s levels nor of the observations within the level.
In this situation, it is possible to obtain a test statistic by first summing scores within
each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where the first subscript on the two terms in parentheses are $i_\ell$
and $i_{\ell-1}$, respectively (and where we take $i_0 = 0$, so that
the cumulative score is ${\bm B}(0, \hat {\bm \theta}) = \bm{0}$).
Again, tests of subsets of model parameters can be obtained by 
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and likelihood ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.
In contrast, the statistics proposed above explicitly account for the fact
that $V$ is ordinal.

@@

plot(gefp1, functional = ordwmax(gefp1), axes = FALSE, main = "",
     xlab = "Age group", ylab = "Weighted max statistics",
     boundary = FALSE)
lines(get_boundary(gefp1, ordwmax(gefp1)), col = "slategray", lwd = 1.7)
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)

plot(gefp1, functional = ol2bb1.fun, axes = FALSE, ylim = c(0, 14), main = "",
     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
lines(get_boundary(gefp1, ol2bb1.fun), col = "slategray", lwd = 1.7)
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)
@
\end{figure}

@@

plot(gefp2, functional = ordwmax(gefp2), axes = FALSE, main = "",
     xlab = "Age group", ylab = "Weighted max statistics",
     boundary = FALSE)
lines(get_boundary(gefp2, ordwmax(gefp2)), col = "slategray", lwd = 1.7)
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)

plot(gefp2, functional = ol2bb2.fun, axes = FALSE, ylim = c(0, 115), main = "",
     xlab = "Age group", ylab = "LM statistics", boundary = FALSE)
lines(get_boundary(gefp2, ol2bb2.fun), col = "slategray", lwd = 1.7)
axis(1, at = 1:5, labels = c("10-11", "12-13", "14", "15", "16"))
axis(2)
@
\end{figure}
